---
title: "NIH NASA Project"
author: "Diyanko Bhowmik"
date: "4/22/2021"
output: html_document
---

## Importing Packages

```{r}
library(dplyr)
library(ggplot2)
library(usmap)
library(cluster)
library(fpc)
library(Dict)
library(dendextend)
library(tidyverse)
```

## Importing Data

```{r}
prec <- read.csv("D:/Users/Diyanko Bhowmik/Desktop/Data Analytics/DataAnalytics2021-Diyanko-Bhowmik/Project/data/precipitation_data/prec_for_centers_monthly.csv")
elev <- read.csv("D:/Users/Diyanko Bhowmik/Desktop/Data Analytics/DataAnalytics2021-Diyanko-Bhowmik/Project/data/elevation_data/elevation_for_centers.csv")
centers <- read.csv("D:/Users/Diyanko Bhowmik/Desktop/Data Analytics/DataAnalytics2021-Diyanko-Bhowmik/Project/data/centers_data/combined_centers_data.csv")
pop_density <- read.csv("D:/Users/Diyanko Bhowmik/Desktop/Data Analytics/DataAnalytics2021-Diyanko-Bhowmik/Project/data/population_density/population_density.csv")
```

## EDA

### All Centers

```{r}
centers_location = data.frame(LONGITUDE = centers$LONGITUDE, LATITUDE = centers$LATITUDE)

centers_map <- usmap_transform(centers_location)

plot_usmap(include = c("KY", "TN")) +
  geom_point(data = centers_map, aes(x = LONGITUDE.1, y = LATITUDE.1), color = "red") + labs(title = "Health-care centers", subtitle = "All hospitals and urgent health care facilities in Kentucky and Tenessee") + theme(legend.position = "right")
```

Some important variables

```{r}
IDs = prec[1]
IDs = unlist(IDs)
```

#### Functions

##### Getting precipitation data for a location

```{r}
get_prec = function(ID){
  prec_data = prec[prec$ID == ID,-1]
  temp_matrix = matrix(unlist(prec_data), nrow = 12, byrow = FALSE)
  return(temp_matrix)
}

#, dimnames = list(c(1:12),c(2005:2020)
```

##### Getting Annual precipitation

```{r}
get_annual_prec = function(ID){
  temp_matrix = get_prec(ID)
  return(matrix(colSums(temp_matrix, na.rm = FALSE, dims = 1)))
}
```

##### Getting Average Annual precipitation

```{r}
get_avg_annual_prec = function(ID){
  return(mean(get_annual_prec(ID)))
}
```

##### Getting Average Monthly precipitation

```{r}
get_monthly_prec = function(ID){
  temp_matrix = get_prec(ID)
  return(matrix(rowSums(temp_matrix, na.rm = FALSE, dims = 1))/16)
}
```

##### Normalization

```{r}
normalize <- function(x){
  return ((x-min(x))/(max(x)-min(x)))
}
```

#### Average Annual Precipitation at centers

```{r}
annual_prec_location = apply(prec[-1], 1, sum)

centers_location = data.frame(LONGITUDE = centers$LONGITUDE, LATITUDE = centers$LATITUDE, annual_prec_location)

centers_map <- usmap_transform(centers_location)

annual_prec_location = normalize(annual_prec_location)

plot_usmap(include = c("KY", "TN")) +
  geom_point(data = centers_map, aes(x = LONGITUDE.1, y = LATITUDE.1), size = annual_prec_location*5, color = annual_prec_location+5.4, alpha = 0.3) + 
  labs(title = "Average Annual Precipitation", subtitle = "All hospitals and urgent health care facilities in Kentucky and Tenessee") + 
  theme(legend.position = "right")
```

#### Elevation at centers

```{r}
centers_elevation = data.frame(LONGITUDE = centers$LONGITUDE, LATITUDE = centers$LATITUDE, elev$elevation)

centers_elev <- usmap_transform(centers_elevation)

plot_usmap(include = c("KY", "TN")) +
  geom_point(data = centers_elev, aes(x = LONGITUDE.1, y = LATITUDE.1, color = elev.elevation), size = 5, alpha = 0.25) + 
  scale_color_gradientn(colours = terrain.colors(3), breaks = seq(0, 255, by = 25)) +
  labs(title = "Elevation above sea-level", subtitle = "All hospitals and urgent health care facilities in Kentucky and Tenessee") + 
  theme(legend.position = "right")
```

```{r}
boxplot(centers_location$annual_prec_location*1000, outline=FALSE, xlab = "Average Annual Precipitation (in mm)")
hist(centers_location$annual_prec_location*1000,seq(27000,34000,500), main = paste("Histogram of precipitation distribution"), xlab = "Average Annual Precipitation (in mm)", ylab = "Number of locations")

```

```{r}
#get_monthly_prec(IDs[1,1])

#hello = sapply(IDs, get_monthly_prec, ID)
```

#### Annual rainfall in all centers

```{r}
yearly_prec_data = data.frame(months = c(1:12))

for (id in 1:503) {
  yearly_prec_data = cbind(yearly_prec_data, get_monthly_prec(IDs[id]))
}

yearly_prec_data = yearly_prec_data[,-1]

matplot(yearly_prec_data*1000,  type = c("l"), pch=1, col='black', xlab = "Months", ylab = "Average Monthly Precipitation (in mm)")
```

#### Year on year change in precipitation

This shows the percentage variation in the average monthly rainfall of the last few years compares to the duration before that.

We can see that there has been a big increase in February and October when we compare the data of 2019-2020 to 2005-2018

```{r}
n2 = 3
n1 = 16 - n2
m1 = 1+(n1*12)


get_prec_2 = function(ID,temp_data){
  prec_data = temp_data[prec$ID == ID,-1]
  temp_matrix = matrix(unlist(prec_data), nrow = 12, byrow = FALSE)
  return(temp_matrix)
}

get_monthly_prec_2 = function(ID,temp_data,number){
  temp_matrix = get_prec_2(ID,temp_data)
  return(matrix(rowSums(temp_matrix, na.rm = FALSE, dims = 1))/number)
}

prec_n1 = prec[,1:m1]
prec_n2 = prec[,-(2:m1)]

prec_data_n1 = data.frame(months = c(1:12))
prec_data_n2 = data.frame(months = c(1:12))


for (id in 1:503) {
  prec_data_n1 = cbind(prec_data_n1, get_monthly_prec_2(IDs[id],prec_n1,n1))
  prec_data_n2 = cbind(prec_data_n2, get_monthly_prec_2(IDs[id],prec_n2,n2))
}

prec_data_n1 = prec_data_n1[,-1]
prec_data_n2 = prec_data_n2[,-1]

prec_data_diff = (prec_data_n2 - prec_data_n1)/prec_data_n1*100

matplot(prec_data_diff,  type = c("l"), pch=1, col="black", xlab = "Months", ylab = "% Change in monthly precipitation")
```

#### Year on year change in precipitation wrt location

This is a spatial representation of the change in precipitation for the two states.

```{r}
analysis_month = 2
prec_data_diff_Feb = prec_data_diff[analysis_month,]

prec_data_diff_Feb = unlist(prec_data_diff_Feb)

prec_diff_centers_location = data.frame(LONGITUDE = centers$LONGITUDE, LATITUDE = centers$LATITUDE, prec_data_diff_Feb)

prec_diff_centers_map <- usmap_transform(prec_diff_centers_location)

prec_data_diff_Feb = normalize(prec_data_diff_Feb)

plot_usmap(include = c("KY", "TN")) +
  geom_point(data = prec_diff_centers_map, aes(x = LONGITUDE.1, y = LATITUDE.1), size = prec_data_diff_Feb*5, magnitude = 1, color = (prec_data_diff_Feb+2.45), alpha = 0.5) + 
  labs(title = "Change in Precipitation at centers (on February)", subtitle = "All hospitals and urgent health care facilities in Kentucky and Tenessee") + 
  theme(legend.position = "right")
```

#### Population Density and Change in precipitation

This compares the spatial representation of the change in precipitation for the two states with that of the population density.

```{r}
centers_pop_density = select(centers, ID, LATITUDE, LONGITUDE, ZIP)

centers_pop_density$POP_DENSITY = pop_density$Density.Per.Sq.Mile[match(centers_pop_density$ZIP, pop_density$Zip.ZCTA)]


pop_location = centers_pop_density$POP_DENSITY

centers_pop_density_modified = data.frame(LONGITUDE = centers$LONGITUDE, LATITUDE = centers$LATITUDE, pop_location)

centers_pop_density_map <- usmap_transform(centers_pop_density_modified)

#pop_location = normalize(pop_location)

plot_usmap(include = c("KY", "TN")) +
  geom_point(data = centers_pop_density_map, aes(x = LONGITUDE.1, y = LATITUDE.1), size = pop_location/500, color = "red", alpha = 0.2) + geom_point(data = prec_diff_centers_map, aes(x = LONGITUDE.1, y = LATITUDE.1), size = prec_data_diff_Feb*5, magnitude = 1, color = "blue", alpha = 0.2) + 
  labs(title = "Population Density", subtitle = "All hospitals and urgent health care facilities in Kentucky and Tenessee") + 
  theme(legend.position = "right")
```

## Models

### K-Means

```{r}
set.seed(300)
k.max <- 12
wss <- sapply(1:k.max,function(k){kmeans(centers_location$annual_prec_location,k,nstart = 50,iter.max = 5000)$tot.withinss})

wss

plot(1:k.max,wss, type= "b", xlab = "Number of clusters(k)", ylab = "Within cluster sum of squares")

icluster <- kmeans(centers_location$annual_prec_location,4,nstart = 50)

plotcluster(centers_location$annual_prec_location*1000, icluster$cluster, xlab = "Annual Precipitation (in mm)", ylab = "Cluster Number")

icluster$centers
table(icluster$cluster)

plot_usmap(include = c("KY", "TN")) +
  geom_point(data = prec_diff_centers_map, aes(x = LONGITUDE.1, y = LATITUDE.1), size = 4, magnitude = 1, color = icluster$cluster, alpha = 0.5) + 
  labs(title = "Clusters based on Precipitation", subtitle = "All hospitals and urgent health care facilities in Kentucky and Tenessee") + 
  theme(legend.position = "right")

kmeans_clustered = data.frame(centers_location, K=icluster$cluster)
```

### Agglomerative Hierarchical Clustering

```{r}
annual_prec_location_scaled = as.data.frame(scale(centers_location$annual_prec_location))

dist_mat <- dist(annual_prec_location_scaled, method = 'euclidean')

hclust_avg <- hclust(dist_mat, method = 'complete')
plot(hclust_avg, cex = 0.6, hang = -1)

cut_avg <- cutree(hclust_avg, k = 4)
rect.hclust(hclust_avg , k = 4, border = 2:5)

suppressPackageStartupMessages(library(dendextend))

dendogram = as.dendrogram(hclust_avg)
dendogram_color = color_branches(dendogram, k=4)
plot(dendogram_color)

aggl_hier_clustered <- mutate(centers_location, cluster = cut_avg)

table(aggl_hier_clustered$cluster)

plot_usmap(include = c("KY", "TN")) +
  geom_point(data = prec_diff_centers_map, aes(x = LONGITUDE.1, y = LATITUDE.1), size = 4, magnitude = 1, color = aggl_hier_clustered$cluster, alpha = 0.5) + 
  labs(title = "Clusters based on Precipitation", subtitle = "All hospitals and urgent health care facilities in Kentucky and Tenessee") + 
  theme(legend.position = "right")
```

### Comparison

```{r}
kmeans_clustered$level[kmeans_clustered$K == 3] = 1
kmeans_clustered$level[kmeans_clustered$K == 4] = 2
kmeans_clustered$level[kmeans_clustered$K == 2] = 3
kmeans_clustered$level[kmeans_clustered$K == 1] = 4

aggl_hier_clustered$level[aggl_hier_clustered$cluster == 4] = 1
aggl_hier_clustered$level[aggl_hier_clustered$cluster == 3] = 2
aggl_hier_clustered$level[aggl_hier_clustered$cluster == 2] = 3
aggl_hier_clustered$level[aggl_hier_clustered$cluster == 1] = 4


comparing = data.frame(km = kmeans_clustered$level,ahc = aggl_hier_clustered$level)
comparing$diff = ifelse(comparing$km == comparing$ahc, 0,1)

plotcluster(centers_location$annual_prec_location, kmeans_clustered$level)
plotcluster(centers_location$annual_prec_location, aggl_hier_clustered$level)
```
